---
title: "EDA of Energy Usage Data"
output: 
  html_notebook: 
    toc: yes
  html_document: 
    keep_md: yes
    theme: yeti
    toc: yes
---

```{r,echo=FALSE}

knitr::opts_chunk$set(fig.width=12, fig.height=10,echo=TRUE, warning=FALSE, message=FALSE)

```
      


```{r libraries}

library(ggplot2)
theme_set(theme_gray(base_size = 18))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))

```

### Gas data
```{r}
# load gas data
gas <- read.csv('data/xcel_gas.csv',stringsAsFactors = FALSE)
gas <-gas %>% rename(Date=Last.Read.Date,therms=Gas.Usage..Therms.,avgTemp=Average.Temperature) %>% 
  select(Date,therms,avgTemp,Heating.Degree.Days) %>% 
  mutate(Date=mdy(Date)) %>% 
  filter(!is.na(Date)) %>%
  mutate(Yr=as.factor(lubridate::year(Date))) %>%
  mutate(avgTemp=as.integer(gsub("[^0-9]", "", avgTemp) ))
head(gas)

gas$HDD <- gas$Heating.Degree.Days
gas <- gas %>% select(-c(Heating.Degree.Days))

```



### Electric Data
```{r}

ele <- read.csv('data/xcel_elec.csv',stringsAsFactors = FALSE)

ele <- ele %>% rename(Date=Last.Read.Date,kWh=Electric.Usage..kWh.,avgTemp=Average.Temperature) %>% 
  select(Date,kWh,avgTemp,Heating.Degree.Days,Cooling.Degree.Days) %>% 
  mutate(Date=mdy(Date)) %>% 
  filter(!is.na(Date)) %>%
  mutate(Yr=as.factor(lubridate::year(Date))) %>%
  mutate(avgTemp=as.integer(gsub("[^0-9]", "", avgTemp) ))

ele$CDD<-ele$Cooling.Degree.Days
ele$HDD <- ele$Heating.Degree.Days
ele <- ele %>% select(-c(Heating.Degree.Days,Cooling.Degree.Days))
head(ele)

```



# Analysis


### Timeseries 
First i'll look at timeseries of temperature, gas use, and electricity use
```{r}
p1 <- gas %>% ggplot(aes(Date,avgTemp))+
    geom_line(size=1)+
  geom_point(size=3)+
  ggtitle("Average monthly temperature")

p2 <-gas %>% ggplot(aes(Date,therms))+
  geom_line(size=1)+
  geom_point(size=3)+
  ggtitle("Natural Gas Use")

p3 <-ele %>% ggplot(aes(Date,kWh))+
  geom_line(size=1)+
  geom_point(size=3)+
  ggtitle("Electrcity Use")

gridExtra::grid.arrange(p1,p2,p3)

```

### Energy use vs HDD and CDD
```{r}
gas %>% ggplot(aes(HDD,therms))+geom_point(size=4,aes(col=Yr))+geom_smooth(method='lm')
```


```{r}

ele %>% ggplot(aes(HDD,kWh))+geom_point(size=4,aes(col=Yr))+geom_smooth(method='lm')
#ele %>% ggplot(aes(CDD,kWh))+geom_point(size=4,aes(col=Yr))+geom_smooth(method='lm')
```



### Energy use vs temperature
* A scatter plot is good for showing the relationship between energy use and temperature.
* There is a very strong correlation between gas use and temperature
* This makes sense because we have gas heat, and the only other gas appliance we have is the water heater.
* You can see there is sort of a 'break' in the relationship for temperatures above 70 deg. We never use our heat if it's hotter than 70 deg, so gas usage shouldn't depend on temperature above 70 deg.


```{r}


p1 <- gas %>% ggplot(aes(avgTemp,therms))+
  geom_point(size=5,aes(col=Yr))+
  geom_smooth(method='lm')+
  ggtitle("Gas Usage vs Temperature ; All Data")+
   xlab("Monthly Avg. Temperature")

gas2 <- gas %>% filter(avgTemp<70)
p2 <-  gas2 %>% 
  ggplot(aes(avgTemp,therms))+
  geom_point(size=5,aes(col=Yr))+
  geom_smooth(method='lm')+
  ggtitle("Gas Usage vs Temperature ; Only Temp <70")+
   xlab("Monthly Avg. Temperature")


gridExtra::grid.arrange(p1,p2)

```

### Linear regression vs Temperature

We can fit a linear regression to quantify the dependence of gas usage on temperature. I fit 2 models; 1 with all the data, and another excluding data where temperature>70.

#### Model w/ All Data
```{r}
model1 <- lm(therms~avgTemp,data=gas)
broom::tidy(model1)
```

#### Model w/ Only Temperature<70
```{r}
model2 <- lm(therms~avgTemp,data=gas %>% filter(avgTemp<70))
broom::tidy(model2)
```

* Both models return estimates that are statistically significant (very small p-values). I'll use the model excluding temperatures <70, since we definitely don't use any gas heat in that range.
* The model slope is -2.71, which means that for every degree colder it is (below 70 degress), we use 2.71 more therms of natural gas. 
* We are planning to add insulation to our attic, so it will be interesting to see if the slope changes after that (hopefully it does!)





### Electricity use vs Temperature

* From the 1st timeseries plot, the electricity usage looks fairly constant, except for a spike in January 2017. We have gas heat, so *why the spike*?
* Well, when we bought the house there was no heat vent in the back room and we used an electric heater there; the spike in January is likely due to that. We had a vent installed in February and didn't need to use the electric heater anymore. 
* The first value (Sept. 2016) is also lower than the rest; probably because we were moving in during this period and our first bill didn't cover a full month.

```{r}


p1 <- ele %>% ggplot(aes(avgTemp,kWh))+
  geom_point(size=5, aes(col=Yr) )+
  geom_smooth(method = 'lm')+
  ggtitle("kWh vs. Temperature : All Data")+
   xlab("Monthly Avg. Temperature")


p2 <- ele %>% filter(kWh<450 & kWh>200) %>%
  ggplot(aes(avgTemp,kWh))+
  geom_point(size=5, aes(col=Yr) )+
  geom_smooth(method = 'lm') +
  ggtitle("kWh vs. Temperature : Sept 2016 and January 2017 Removed")+
 xlab("Monthly Avg. Temperature")
   
gridExtra::grid.arrange(p1,p2)


```

### Linear Regression vs. temperature
A linear regression of electricty use vs temperature shows no significant relationship (the slope pvalue is 0.5).
```{r}
model_ele <- lm(kWh~avgTemp,data = ele %>% filter(kWh<450 & kWh>200) )
broom::tidy(model_ele)
```




* The scatter plots above show there isn't really much of a relationship between electricity use and temperature. The 3 higher points above 70 degrees might be due to our window AC unit, but other than that there isn't much seasonality to our electricity use.

* We also just got an electric car that we charge at home every few nights, so it will be interesting to see how that changes our electricity usage.


## Try it out with your energy data and let me know what you find!